Supporting Information for\\ ``A weakly structured stem for human origins in Africa''

### 2022-02-19 supplement.tex
clean up alpha description

To evaluate the degree to which divergence among modern populations is
inherited from divergence between early stem population, we computed $f_4$
statistics of the form
\begin{equation}
    f_4(\text{stem}_1, \text{stem}_2; \text{modern}_1, \text{modern}_2) = 
    \E[(p_{\text{stem}_1} - p_{\text{stem}_2} )(p_{\text{modern}_1} -p_{\text{modern}_2} )]
    \label{eq:f4}
\end{equation}
This statistic measures co-linearity of genetic drift between modern
populations $(p_{\text{modern}_1} -p_{\text{modern}_2} )$ and stem populations
$(p_{\text{stem}_1} -p_{\text{stem}_2} )$ (see
Figures~\ref{fig:supp-f4s-single-origin}--\ref{fig:supp-f4s-merger-with-stem-migration}
and \citet{Lipson2020-cq}).

To contrast these statistics to divergence across modern populations,
\begin{equation}
    f_2(\text{modern}_1, \text{modern}_2) =
    \E[(p_{\text{modern}_1} -p_{\text{modern}_2} )^2]),
\end{equation}
we interpret the $f$-statistics as scalar products between vectors in 
allele frequency space. In particular, the vector 
$v=(p_{\text{modern}_1} -p_{\text{modern}_2} )$, 
representing differences in allele frequencies across modern populations,
can be written as $v=v_\perp + v_\parallel$, a component perpendicular and
a component parallel to the divergence $(p_{\text{stem}_1} -p_{\text{stem}_2} )$
across stems. We can then decompose 
$f_2(\text{modern}_1,\text{modern}_2) = v_\perp^2+v_\parallel^2$, with  
\begin{equation}
    v_\perp^2 = f_4^2(\text{stem}_1, \text{stem}_2; 
    \text{modern}_1, \text{modern}_2)/f_2(\text{stem}_1, \text{stem}_2)
\end{equation}

Finally, the proportion of modern-day structure
$f_2(\text{modern}_1, \text{modern}_2)$ that is accounted for by the
component parallel to ancient population structure is
\begin{equation}
    \alpha^2 = \frac{v_\perp^2}{f_2(\text{modern}_1, \text{modern}_2)} =
    \frac{f_4^2(\text{stem}_1, \text{stem}_2; \text{modern}_1, \text{modern}_2)}
    {f_2(\text{stem}_1,\text{stem}_2)f_2(\text{modern}_1, \text{modern}_2)}.
\end{equation}
$\alpha^2$ can be interpreted as the proportion of the $f_2$
divergence across modern populations that is aligned with the differences in
allele frequencies across ancient populations. However, drift occurring after
the time of sampling of ancient populations is not correlated to drift
occurring before that time. For this reason, we also interpret $\alpha^2$ as the
proportion of the $f_2$ divergence across modern populations that is explained by
the differences in allele frequencies across ancient populations. We report
this predicted statistic for each of the four inferred demographic models in
Figures~\ref{fig:supp-f4s-single-origin}--\ref{fig:supp-f4s-merger-with-stem-migration}.

>>>>

To evaluate the degree to which divergence among modern populations is
inherited from divergence between early stem population, we computed $f_4$
statistics of the form
\begin{equation}
    f_4(\text{stem}_1, \text{stem}_2; \text{modern}_1, \text{modern}_2) = 
    \E[(p_{\text{stem}_1} - p_{\text{stem}_2} )(p_{\text{modern}_1} -p_{\text{modern}_2} )]
    \label{eq:f4}
\end{equation}
This statistic measures co-linearity of genetic drift between modern
populations $v= (p_{\text{modern}_1} -p_{\text{modern}_2} )$ and stem populations
$w= (p_{\text{stem}_1} -p_{\text{stem}_2} )$ (see
Figures~\ref{fig:supp-f4s-single-origin}--\ref{fig:supp-f4s-merger-with-stem-migration}
and \citet{Lipson2020-cq} for an overview of $f_4$ statistics).

To contrast these statistics to divergence across modern populations,
\begin{equation}
    f_2(\text{modern}_1, \text{modern}_2) =
    \E[(p_{\text{modern}_1} -p_{\text{modern}_2} )^2],
\end{equation}
we interpret the $f$-statistics as scalar products between vectors $v$ and $w$, i.e.,  
$f_2(\text{modern}_1, \text{modern}_2) = v\cdot v$ and 
$f_4(\text{stem}_1, \text{stem}_2; \text{modern}_1, \text{modern}_2) = v\cdot w.$
We can then write $v=v_\perp + v_\parallel$ as the sum of a component $v_\perp$ perpendicular and
a component $v_\parallel$ parallel to $w = (p_{\text{stem}_1} -p_{\text{stem}_2} ).$
We can then decompose 
$f_2(\text{modern}_1,\text{modern}_2) = v_\perp^2+v_\parallel^2$, with  
\begin{equation}
    v_\parallel^2 = \frac{(v \cdot w)^2}{w\cdot w} = \frac{ f_4^2(\text{stem}_1, \text{stem}_2; 
    \text{modern}_1, \text{modern}_2)}{f_2(\text{stem}_1, \text{stem}_2)}
\end{equation}

Finally, the proportion of modern-day structure
$v^2 = f_2(\text{modern}_1, \text{modern}_2)$ that is accounted for by the
component parallel to ancient population structure is
\begin{equation}
    \alpha^2 = \frac{v_\parallel^2}{v^2} =
    \frac{f_4^2(\text{stem}_1, \text{stem}_2; \text{modern}_1, \text{modern}_2)}
    {f_2(\text{stem}_1,\text{stem}_2)f_2(\text{modern}_1, \text{modern}_2)}.
\end{equation}
$\alpha^2$ can be interpreted as the proportion of the $f_2$
divergence across modern populations that is aligned with the differences in
allele frequencies across ancient populations. However, drift occurring after
the time of sampling of ancient populations is not correlated to drift
occurring before that time. For this reason, we also interpret $\alpha^2$ as the
proportion of the $f_2$ divergence across modern populations that is \emph{explained by}
the differences in allele frequencies across ancient populations. We report
this predicted statistic for each of the four inferred demographic models in
Figures~\ref{fig:supp-f4s-single-origin}--\ref{fig:supp-f4s-merger-with-stem-migration}.
